---
title: 'Basic Course on **R**: The apply family of functions - Practical'
author: "David Nieuwenhuijse"
date: "May 18^th^ - 24^th^, 2017"
output:
  pdf_document:
    number_sections: no
    toc: no
    toc_depth: 3
    fig_height: 4
  html_document:
    toc: yes
    toc_depth: '3'
---

## Part D: Using the apply family of functions

1. Use apply() to turn the following code into something shorter:

```{r}

#This function determines if a number is a prime number
isPrime <- function(num){
  if (num == 2) {
    return(TRUE)
  }
  if(num > 1) {
    for(i in 2:(num-1)) {
      if ((num %% i) == 0) {
        return(FALSE)
      }
    }
  } else {
    return(FALSE)
  }
  
  return(TRUE)
}

#The matrix with numbers to be checked:
mat <- matrix(1:100, nrow=10)
#The matrix with answers (TRUE/FALSE)
answer <- matrix(rep(x = TRUE,100), nrow=10)

for (x in 1:10) {
  for (y in 1:10) {
    answer[x,y] <- isPrime(mat[x,y])
  }
}

#The resulting prime numbers:
mat[answer]

```

Answer:

```{r}

#This function determines if a number is a prime number
isPrime <- function(num){
  if (num == 2) {
    return(TRUE)
  }
  if(num > 1) {
    for(i in 2:(num-1)) {
      if ((num %% i) == 0) {
        return(FALSE)
      }
    }
  } else {
    return(FALSE)
  }
  
  return(TRUE)
}

#The matrix with numbers to be checked:
mat <- matrix(1:100, nrow=10)

#With apply we do not need to create an answer matrix anymore
answer <- apply(mat, c(1,2), isPrime)

#The resulting prime numbers:
mat[answer]

```

2. Answer question 2.3 again using the lapply() function.

```{r}
source("kennedys.txt")

lapply(Kennedys, length)

```

3. Answer question 2.3 again using the sapply() function. What is the class of the output?

```{r}
sapply(Kennedys, length)
class(sapply(Kennedys, length))
```

4. Read in the diamonds.txt dataset using read.table, make sure the headers are correctly loaded. Calculate the average price of diamonds by color and clarity using the tapply function.

```{r}

diamonds <- read.table("diamonds.txt", header = TRUE)

average_price <- tapply(diamonds$Price, list(diamonds$Color, diamonds$Clarity), mean)

print(average_price)

```

